---
title: "Analysis of the ToothGrowth dataset"
author: "Cristian Bidea"
date: "10 April 2016"
output: pdf_document
---

We set ourselves to analize the ToothGrowth dataset and see if we can come up
with some insights.

## Exploratory analysis

```{r, echo=FALSE}
data("ToothGrowth")
summary(ToothGrowth)
```

First we do a summary of the dataset and we can see that the dataset is quite
simple: just three columns. From the help file we find out more information
about this particular dataset.

A data frame with 60 observations on 3 variables.

[,1]	len	numeric	Tooth length
[,2]	supp	factor	Supplement type (VC or OJ).
[,3]	dose	numeric	Dose in milligrams/day

The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, (orange juice or ascorbic acid (a form of vitamin C and coded as VC).

We are now looking at two aspects:

1. First we want to know if the doses administered have different effects in each group (VC, OJ)
2. Second if one method is better than the other on the same administered dose.

As you can see in Fig. 1, in both methods increasing the dose increases the result
and it seems that administering OJ is better for doses of 0.5 and 1 but then it 
becomes a little bit better for a dose of 2.

What I'm saying now is not rigorously tested. These are just intuition based
conclusions formed looking at a graphic. We'll move on in the next section and we'll
try to asses if these differences are in fact meaningful or not.

## Confidence intervals and hypothesis testing

Why is computing confidence intervals useful? We plotted the mean of the samples depending on the
method and dose but we must be confident that we actually have an increase. So 
we are computing confidence interval comparing the means of two independent
populations: first comparing the 0.5 dose group with the 1 dose group and then 
the 1 dose group with the 2 dose group, in each of the two methods.


```{r, echo=FALSE}
require(ggplot2)
mean_tg <- data.frame(
    len=c(0, 0, 0, 0, 0, 0),
    supp=factor(c("VC", "VC", "VC", "OJ", "OJ", "OJ")),
    dose=factor(c(0.5, 1.0, 2.0, 0.5, 1.0, 2.0)),
    sd=c(0, 0, 0, 0, 0, 0),
    ci1=c(NA, NA, NA, NA, NA, NA),
    ci2=c(NA, NA, NA, NA, NA, NA),
    power=c(NA, NA, NA, NA, NA, NA)
)
transform (mean_tg, ci1 = as.numeric(ci1), ci2 = as.numeric(ci2))

tg <- ToothGrowth
lvl <- "VC"
mean_tg[mean_tg$supp == lvl & mean_tg$dose == .5,]$len <- mean(tg[tg$supp == lvl & tg$dose ==.5,]$len)
mean_tg[mean_tg$supp == lvl & mean_tg$dose == 1,]$len <- mean(tg[tg$supp == lvl & tg$dose ==1,]$len)
mean_tg[mean_tg$supp == lvl & mean_tg$dose == 2,]$len <- mean(tg[tg$supp == lvl & tg$dose ==2,]$len)
lvl <- "OJ"
mean_tg[mean_tg$supp == lvl & mean_tg$dose == .5,]$len <- mean(tg[tg$supp == lvl & tg$dose ==.5,]$len)
mean_tg[mean_tg$supp == lvl & mean_tg$dose == 1,]$len <- mean(tg[tg$supp == lvl & tg$dose ==1,]$len)
mean_tg[mean_tg$supp == lvl & mean_tg$dose == 2,]$len <- mean(tg[tg$supp == lvl & tg$dose ==2,]$len)

ggplot(data=mean_tg, aes(x=dose, y=len, fill=supp)) +
        geom_bar(stat="identity", position=position_dodge()) +
        ylab("Mean tooth length milligrams/day") +
        xlab("Dose level") +
        ggtitle("Fig 1. Comparison of mean tooth lengths depending on dose and method") +
        scale_fill_hue(name="Method")
```

We need the standard deviation but we'll infer that from the CLT, knowing that 
```{r}
# computing the standard deviations
lvl <- "VC"
mean_tg[mean_tg$supp == lvl & mean_tg$dose == .5,]$sd <- sd(tg[tg$supp == lvl & tg$dose ==.5,]$len)
mean_tg[mean_tg$supp == lvl & mean_tg$dose == 1,]$sd <- sd(tg[tg$supp == lvl & tg$dose ==1,]$len)
mean_tg[mean_tg$supp == lvl & mean_tg$dose == 2,]$sd <- sd(tg[tg$supp == lvl & tg$dose ==2,]$len)
lvl <- "OJ"
mean_tg[mean_tg$supp == lvl & mean_tg$dose == .5,]$sd <- sd(tg[tg$supp == lvl & tg$dose ==.5,]$len)
mean_tg[mean_tg$supp == lvl & mean_tg$dose == 1,]$sd <- sd(tg[tg$supp == lvl & tg$dose ==1,]$len)
mean_tg[mean_tg$supp == lvl & mean_tg$dose == 2,]$sd <- sd(tg[tg$supp == lvl & tg$dose ==2,]$len)

for (x in c(1:2, 4:5)) 
{
    s1 <- mean_tg[x,]$sd
    s2 <- mean_tg[(x+1),]$sd
    mu1 <- mean_tg[x,]$len
    mu2 <- mean_tg[(x+1),]$len
    sd_p <- (9 * s1^2 + 9 * s2^2) / 18
    se <- sqrt(2*sd_p/10)
    mean_tg$ci1[x] <- mu2 - mu1 - qt(.95, 18) * se
    mean_tg$ci2[x] <- mu2 - mu1 + qt(.95, 18) * se
    mean_tg$power[x] <-power.t.test(n=20, delta = mu2 - mu1, sd = sd_p, type="one.sample", alt="one.sided")$power
}

ci <- data.frame(
    ci1 = mean_tg$ci1[!is.na(mean_tg$ci1)],
    ci2 = mean_tg$ci2[!is.na(mean_tg$ci2)],
    method = mean_tg$supp[!is.na(mean_tg$ci1)],
    power = mean_tg$power[!is.na(mean_tg$ci1)]
)

ci
```

Looking at the confidence intervals we can see that there is an increase in effects
that goes with the increase in dose because none of the four CIs don't touch 0. 

Looking at the graphic in Fig. 1 and then observing that between 1 and 2 the
increase in effects for the OJ method isn't that big, we ask ourselves if it wouldn't
be better to take a lower dose (1) of OJ instead of taking a bigger dose (2) of VC.

So we compute one more confidence level comparing the means of these two groups
to see if there is really an increase for a dose 2 of VC over a dose 1 of OJ.

```{r, echo=FALSE}
    s1 <- mean_tg[5,]$sd
    s2 <- mean_tg[(3),]$sd
    mu1 <- mean_tg[5,]$len
    mu2 <- mean_tg[(3),]$len
    sd_p <- (9 * s1^2 + 9 * s2^2) / 18
    se <- sqrt(2*sd_p/10)
    ci1 <- mu2 - mu1 - qt(.95, 18) * se
    ci2 <- mu2 - mu1 + qt(.95, 18) * se
    power <-power.t.test(n=20, delta = mu2 - mu1, sd = sd_p, type="one.sample", alt="one.sided")$power
    result <- data.frame(
        ci1 = ci1,
        ci2 = ci2,
        power = power
    )
    result
```

## Conclusions

We've observed that there is a real increase of the effects that goes with an
increased dose, especially for the VC group. For the OJ group the increase from
dose 1 to dose 2 is questionable considering the low power.

A surprising result is conveyed by the confidence interval between the means of 
OJ(1) and VC(2) because this shows that the increase isn't that big and the low
power showed us that a bigger sample is given to confirm that.

So it looks like the best bet is on OJ with a dose of 1, because the effects are
bigger than in the case of VC with a dose of 1, and without a powerful increase
in both VC and OJ with a dose of 2.