---
title: "Getting Info out of GoodReads"
author: "Cristian Bidea"
date: "5 August 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I'm parsing the reading challenge page to get info about the books I've read this year.

```{r}
text <- readLines("goodreads_reading_challenge.html", encoding = "UTF-8")
text <- text[grep("<img.*class=\"bookCover\"", text)]
regexpression <- "<a id=\"(.*)\" class=\"(.*)\" href=\"(.*)\"><img alt=\"(.*)\" class=\"(.*)\" src=\"(.*)\"></a>"
titles <- lapply(text, function(x) {
    m <- regexec(regexpression, x)
    regm <- regmatches(x, m)
    result <- regm[[1]][5]
})

links <- lapply(text, function(x) {
    m <- regexec(regexpression, x)
    regm <- regmatches(x, m)
    result <- regm[[1]][4]
})

book_ids <- lapply(text, function(x) {
    m <- regexec(regexpression, x)
    regm <- regmatches(x, m)
    result <- sub("./goodreads_reading_challenge_files/", "", regm[[1]][7])
    result <- sub(".jpg", "", result)
})

books <- data.frame(Title=simplify2array(titles),
                    Link=simplify2array(links),
                    BookID=simplify2array(book_ids))
```

## Get detailed info about the books

The goal is to get start_date, end_date, rating and number of pages and compile
something resembling [Goodreads Year in review for 2015](https://www.goodreads.com/user/year_in_books/2015).

Average rating and number of pages, we can get them using the Good Reads API,
needing only a developer key.

```{r}
library(RCurl)
library(XML)

if (!exists("goodreads_devkey"))
{
    stop("please define goodreads_devkey and assign it your goodreads devkey")
}

books$AverageRating <- 0
books$NumPages <- 0
i <- 0

rs <- lapply(books$BookID, function(x) {
    address <- paste("https://www.goodreads.com/book/show?id=", x, "&format=xml&key=", goodreads_devkey, sep="")
    print(address)
    xml <- getURL(address)
    xml <- simplify2array(strsplit(xml, "\n"))
    avr <- xml[grep("average_rating", xml)][1]
    avr <- sub("<average_rating>", "", avr)
    avr <- sub("</average_rating>", "", avr)
    
    numpages <- xml[grep("num_pages", xml)][1]
    numpages <- sub("<num_pages><!.CDATA.", "", numpages)
    numpages <- sub("..></num_pages>", "", numpages)
    i <<- i + 1
    
    books$AverageRating[i] <<- as.numeric(avr)
    books$NumPages[i] <<- as.numeric(numpages)
    
    # save the data for later
    write.csv(books, file="books.csv")
})
```

##Produce the infografic

Read the data from *books.csv* and make the Infographic.

```{r}
    library(jpeg)
    library(grid)
    library(gtable)
    library(gridExtra)
    
    books <- read.csv("books.csv")
    books$NumPages[is.na(books$NumPages)] <- 39 # manual value
    
    books$Title <- factor(books$Title, levels=books$Title[order(books$NumPages)])  
    
    mp <- qplot(1:14, 1:14, geom="blank")
    i <- 0.5
    covers <- lapply(books$BookID[order(books$NumPages)], function(x) {
        img <- readJPEG(paste("pics/", x, ".jpg", sep=""))
        cover <- rasterGrob(img, interpolate=TRUE)
        mp <<- mp + annotation_custom(cover, xmin=i, xmax=(i+1))
        i <<- i + 1
        return(cover)
    })
    mp <- mp + geom_blank()
    mp <- mp + theme_bw()
    mp <- mp + theme(axis.title.x = element_blank(), 
                     axis.title.y = element_blank(),
                     axis.text.x = element_blank(),
                     axis.text.y = element_blank(),
                     axis.ticks.x = element_blank(),
                     axis.ticks.y = element_blank(),
                     panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(),
                     panel.border = element_blank(),
                     panel.background = element_blank())

    p <- ggplot(data=books, aes(x=Title, y=NumPages)) +
        geom_bar(colour="black", fill="#FFDDCC", stat="identity") +
        theme_classic() +
        theme(axis.text.x = element_text(angle=90, hjust = 1)) +
        annotate("text", label=paste("Total number of pages:", sum(books$NumPages)), x = 3, y = 500) +
        annotate("text", label=paste("Median number of pages: ", median(books$NumPages)), x = 3, y = 400) +
        geom_hline(yintercept = median(books$NumPages)) +
        theme(axis.title.x = element_blank(),
              axis.title.y = element_blank())
        
    g <- ggplotGrob(p)
    g <- gtable_add_rows(g, unit(4, "cm"), pos=0)
    g <- gtable_add_grob(g, ggplotGrob(mp), 1, 4)
    grid.newpage()
    grid.draw(g)
```